---
title: "individual1"
author: "cyber-atlas"
date: "February 19, 2019"
output: html_document
---

1)
```{r}
library("ggplot2") 
library("dplyr")

#data(happy, package="productplots")
#head(happy)

#wtsall is the weight?
HAPPY <- readRDS("HAPPY.rds")
#head(HAPPY)
#Checks through the dataframe and looks for what satisfies the condition
HAPPY[HAPPY == "IAP" | HAPPY == "DK" | HAPPY == "NA"] <- NA
head(HAPPY)
``` 
2) 

```{r}
class(HAPPY)
sapply(HAPPY, class)

#?as.numeric
#?as.character

#Sets all ages 89+ to 89 and saves as numeric
HAPPY$AGE[HAPPY$AGE == '89 OR OLDER'] <- 89
HAPPY$AGE <- as.numeric(HAPPY$AGE)
#str(HAPPY)


#Use factor to assign levels
HAPPY <- HAPPY %>% mutate(HAPPY = factor(HAPPY, levels = c('NOT TOO HAPPY', 'PRETTY HAPPY', 'VERY HAPPY'))) 
HAPPY <- HAPPY %>% mutate(SEX = factor(SEX, levels = c("FEMALE", "MALE")))
#Marital order levels according to avg age
#can't have non existant age
HAPPY <- na.omit(HAPPY)
maritalVector <- aggregate(AGE ~ MARITAL, data=HAPPY, mean) %>% arrange(AGE)
maritalVector <- as.vector(maritalVector$MARITAL)
HAPPY <- HAPPY %>% mutate(MARITAL = factor(MARITAL, maritalVector))

HAPPY <- HAPPY %>% mutate(DEGREE = factor(DEGREE, levels = c("LT HIGH SCHOOL", "HIGH SCHOOL", "JUNIOR COLLEGE","BACHELOR","GRADUATE")))
HAPPY <- HAPPY %>% mutate(FINRELA = factor(FINRELA, levels = c("FAR BELOW AVERAGE", "BELOW AVERAGE", "AVERAGE", "ABOVE AVERAGE", "FAR ABOVE AVERAGE"))) 
HAPPY <- HAPPY %>% mutate(HEALTH = factor(HEALTH, levels = c("POOR", "FAIR", "GOOD", "EXCELLENT")))
HAPPY <- HAPPY %>% mutate(PARTYID = factor(PARTYID, levels = c("STRONG DEMOCRAT", "NOT STR DEMOCRAT", "IND,NEAR DEM", "INDEPENDENT", "IND,NEAR REPUBLICAN", "NOT STR REPUBLICAN", "OTHER PARTY")))
HAPPY <- HAPPY %>% mutate(POLVIEWS = factor(POLVIEWS, levels = c("EXTREMELY LIBERAL",  "LIBERAL", "SLIGHTLY LIBERAL","MODERATE",  "SLIGHTLY CONSERVATIVE","CONSERVATIVE", "EXTRMLY CONSERVATIVE")))


#mutate factor, says to assign levels in following order

#sapply(HAPPY, class)

#TODO why order factors in each column of a dataframe
#can write out column name instead of for loop

#the graphs are arbitrary

```


1
```{r}


ggplot(data= HAPPY, aes(x=AGE))+
  geom_histogram(aes(color="red"), binwidth = 2)

```


Vased on looking at the distribution of the ages, we can see that a majority of the respondants are young (in 20s and 30s)

2. 
```{r}

ggplot(data = HAPPY, aes(x = HAPPY)) +
  geom_bar(aes(color='red', fill=SEX)) + 
  facet_wrap(HAPPY$MARITAL)

```

Based on the graph it shows that as a whole females are happier than males on average. Both men and women are pretty happy or happy during marriage. Seperated and widowed people are the least happy. 



```{r}

ggplot(data = HAPPY, aes(x = HAPPY)) +
  geom_bar(aes(fill=SEX)) + 
  facet_wrap(HAPPY$FINRELA)

```


I learned that people are happiest when they are at average financial status. Not when rich